### 2023-10-25 calculus.tex
Fix more typos and address some of the feedback

The marked lambda calculus is based on the gradually typed lambda calculus \cite{Siek06a} extended
with numbers, booleans, pairs, and empty expression holes \cite{HazelnutSNAPL}.
Given in \cref{fig:calculus-syntax}, the syntax consists of two expression languages:
%
\begin{itemize}
  \item The \textbf{unmarked language}, which is the original language. Expressions of this language
    are called \emph{unmarked expressions}, denoted by the metavariable $\EMV$.

  \item The \textbf{marked language}, which mirrors the structure of the unmarked language but is
    extended with explicit \emph{error marks}. We call expressions of this language
    \emph{marked expressions}, denoted $\ECMV$.
\end{itemize}
In this simple setting, we only need one sort for types, $\tau$. 
The base types $\TNum$ and $\TBool$ classify number and boolean
expressions. The number literal corresponding to the mathematical number $n$ is given by $\ENumMV$,
and there is a single arithmetic operation, addition. $\ETrue$ and
$\EFalse$ are the boolean values and
$\EIf{\EMV_1}{\EMV_2}{\EMV_3}$ is the boolean conditional. Arrow and product types classify lambda abstractions and pairs, respectively, in the usual way.

$\TUnknown$ is a type hole, which we identify with the unknown type from gradual type theory \cite{Siek06a}.
Finally, to model the edit state of a program in development, $\EEHole$ denotes an \emph{empty expression hole},
used to represent syntactically incomplete portions of the program à la Hazelnut
\cite{HazelnutPOPL}. Note, however, that empty expression holes are not semantically critical to the calculus---they are included to emphasize that the system can handle program sketches.

@@

Finally, the key judgement, \emph{marking}, is also of a bidirectional nature. The synthetic marking judgment
$\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$ establishes that, under the context $\Gamma$, the
unmarked expression $\EMV$ is ``marked into'' the marked expression $\ECMV$, which synthesizes type
$\TMV$. Analogously, the analytic marking judgment $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$
states that $\EMV$ is marked into $\ECMV$, which analyzes against $\TMV$.

@@

  \judgment{
    \notMatchedArrow{\TMV_3} \\
    \ctxAnaTypeM{\extendCtx{\ctx}{x}{\TMV}}{\ECMV}{\TUnknown}
  }{
    \ctxAnaTypeM{\ctx}{\ECLamAnaNonMatchedArrow{x}{\TMV}{\ECMV}}{\TMV_3}
  }{MALam2}
\end{mathpar}
%
Though no expected output type is known, we still need to check and mark the body; it is thus
analyzed against the unknown type in \textsc{MKALam2}.

>>>>

The marked lambda calculus is based on the gradually typed lambda calculus \cite{Siek06a} extended
with numbers, booleans, pairs, and empty expression holes \cite{HazelnutSNAPL}.
Given in \cref{fig:calculus-syntax}, the syntax consists of two expression languages:
%
\begin{itemize}
  \item The \emph{unmarked language}, which is the original language. Expressions of this language
    are called \emph{unmarked expressions}, denoted by the metavariable $\EMV$.

  \item The \emph{marked language}, which mirrors the structure of the unmarked language but is
    extended with explicit \emph{error marks} which, intuitively, may be viewed as exactly the
    errors marked in an editor. We call expressions of this language \emph{marked expressions},
    denoted $\ECMV$.
\end{itemize}
In this simple setting, we only need one sort for types, $\tau$. 
The base types $\TNum$ and $\TBool$ classify number and boolean
expressions. The number literal corresponding to the mathematical number $n$ is given by $\ENumMV$,
and there is a single arithmetic operation, addition. $\ETrue$ and
$\EFalse$ are the boolean values and
$\EIf{\EMV_1}{\EMV_2}{\EMV_3}$ is the boolean conditional. Arrow and product types classify lambda abstractions and pairs, respectively, in the usual way.

$\TUnknown$ is a type hole, which we identify with the unknown type from gradual type theory \cite{Siek06a}.
Finally, to model the edit state of a program in development, $\EEHole$ denotes an \emph{empty expression hole},
used to represent syntactically incomplete portions of the program à la Hazelnut
\cite{HazelnutPOPL}. Note, however, that these empty expression holes are not at all semantically
critical to the calculus---they are included illustrate how the calculus fits into the larger
problem of modelling incomplete program states and to discuss polymorphic generalization in
\cref{sec:polymorphicGlobal}.

@@

Finally, the \emph{marking} judgment is also of bidirectional nature. The synthetic marking judgment
$\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$ establishes that, under the context $\Gamma$, the
unmarked expression $\EMV$ is ``marked into'' the marked expression $\ECMV$, which synthesizes type
$\TMV$. Analogously, the analytic marking judgment $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$
states that $\EMV$ is marked into $\ECMV$, which analyzes against $\TMV$.

@@

  \judgment{
    \notMatchedArrow{\TMV_3} \\
    \ctxAnaTypeM{\extendCtx{\ctx}{x}{\TMV}}{\ECMV}{\TUnknown}
  }{
    \ctxAnaTypeM{\ctx}{\ECLamAnaNonMatchedArrow{x}{\TMV}{\ECMV}}{\TMV_3}
  }{MALam2}
\end{mathpar}
%
Though no expected output type is known, we still need to check and mark the body; it is thus
analyzed against the unknown type in \textsc{MKALam2}. It is possible to instead synthesize the
body, but we choose analysis so that the body is always in analytic mode. The guiding design
principle in this decision is is a notion of ``conservation of mode,'' but it is not critical to the
calculus.

### 2023-10-25 conclusion.tex
Updates from Overleaf
Fix more typos and address some of the feedback

\begin{quote}
    \emph{Nothing will ever be attempted if all possible objections must first be overcome.} 
    \begin{flushright}- Samuel Johnson\end{flushright}
\end{quote}

Programming is increasingly a live collaboration between human programmers and sophisticated semantic services. 
These services need to be able to reason throughout the programming process, not just when the program is formally complete. This paper lays down rigorous type-theoretic foundations for doing just that. Local type inference helps make the localization decisions we make systematic and predictable, and type hole inference shows how local and constraint-based type inference might operate hand-in-hand, rather than as alternatives. 
We hope that language designers will use the techniques introduced in this paper to consider more rigorously, perhaps even formally, the problems of type error localization and error recovery when designing future languages.

>>>>

\begin{quote}
    \emph{Nothing will ever be attempted if all possible objections must first be overcome.} 
    \begin{flushright}-- Samuel Johnson\end{flushright}
\end{quote}

Programming is increasingly a live collaboration between human programmers and sophisticated semantic services. These services need to be able to reason throughout the programming process, not just when the program is formally complete. This paper lays down rigorous type-theoretic foundations for doing just that. Bidirectional type checking helps make the localization decisions we make systematic and predictable, and type hole inference shows how local and constraint-based type inference might operate hand-in-hand, rather than as alternatives. 
We hope that language designers will use the techniques introduced in this paper to consider more rigorously, perhaps even formally, the problems of type error localization and error recovery when designing future languages.

% \section*{Acknowledgements}
% (omitted for review)

% Thank you for reading this paper!

### 2023-10-25 intro.tex
Fix more typos and address some of the feedback

Consider, for example, the ill-typed program below, which is shown as presented to the user in
this paper's version of Hazel \cite{hazel}, a typed functional dialect of Elm \cite{elm}. Hazel  
supports local type inference, specified in the well-established bidirectional style \cite{Localinf,HazelnutPOPL,BidirTyping}:
\vspace{-3px}
\begin{center}
    \includegraphics[scale=0.5]{images/hazel-intro-screenshot-v3.png}
\end{center}
\vspace{-3px}
A type checker with no support for error localization or recovery---%
as students in an undergraduate course might write---%
would simply report that this program is ill-typed. 
A more practical approach, and one common even in production type checkers, 
is to localize the first error that causes local type inference to fail and emit an explanatory error message before terminating.
In this example, the system might report that the variable \li{f} located on Line 2 is free, then terminate.

>>>>

Consider, for example, the ill-typed program below, which is shown as presented to the user in
this paper's version of Hazel \cite{hazel}, a typed functional dialect of Elm \cite{elm}. Hazel  
supports local type inference, specified in the well-established bidirectional style \cite{Localinf,HazelnutPOPL,BidirTyping}:
\\\vspace{-5px}
\begin{center}
    \includegraphics[scale=0.5]{images/hazel-intro-screenshot-v3.png}
\end{center}

A type checker with no support for error localization or recovery---%
as students in an undergraduate course might write---%
would simply report that this program is ill-typed. 
A more practical approach, and one common even in production type checkers, 
is to localize the first error that causes type-checking to fail and emit an explanatory error message before terminating.
In this example, the system might report that the variable \li{f} located on Line 2 is free, then terminate.

### 2023-10-25 polymorphism.tex
Fix non-THI typos from Review C
Fix more typos and address some of the feedback

Likewise, type application mirrors ordinary application: in the case that the expression being applied does not synthesize a matched forall type, an error is marked. With this, totality of marking is satisfied.
%
\begin{mathpar}
  \inferrule[MKSTypeAp2]{
    \bothCtxSynFixedInto{\tvarCtx}{\ctx}{\EMV}{\ECMV}{\MTMV} \\
    \tvarCtxTypeMarkedInto{\tvarCtx}{\TMV_2}{\MTMV_2} \\
    \notMatchedForall{\MTMV}
  }{
    \bothCtxSynFixedInto{\tvarCtx}{\ctx}{\ETypeAp{\EMV}{\TMV_2}}{\ECTypeAp{\ECMV}{\MTMV_2}}{\MTUnknown}
  }

@@

We leave as future work the task of defining marked versions of even more elaborate bidirectional
type systems, e.g. \citet{dunfield2013complete}'s rather substantial formulation of implicit type
application in a bidirectional setting (which would have to handle, for example, the situation where
no implicit argument can be resolved) or \citet{DBLP:phd/hal/LennonBertrand22}'s gradual bidirectional
variant of the dependently typed calculus of constructions.

% TODO cite gradual dependent types papers, etc.
% DBLP:phd/hal/LennonBertrand22

>>>>

Likewise, type application mirrors ordinary application: in the case that the expression being applied does not synthesize a matched forall type, an error is marked. With this, totality of marking is satisfied.
%
\begin{mathpar}
  \inferrule[MKSTypeAp2]{
    \bothCtxSynFixedInto{\tvarCtx}{\ctx}{\EMV}{\ECMV}{\MTMV} \\
    \tvarCtxTypeMarkedInto{\tvarCtx}{\TMV_2}{\MTMV_2} \\
    \notMatchedForall{\MTMV}
  }{
    \bothCtxSynFixedInto{\tvarCtx}{\ctx}{\ETypeAp{\EMV}{\TMV_2}}{\ECTypeApSynNonMatchedForall{\ECMV}{\MTMV_2}}{\MTUnknown}
  }

@@

We leave as future work the task of defining marked versions of even more elaborate bidirectional
type systems, e.g. \citet{dunfield2013complete}'s rather substantial formulation of implicit type
application in a bidirectional setting (which would have to handle, for example, the situation where
no implicit argument can be resolved) or \citet{DBLP:phd/hal/LennonBertrand22}'s gradual bidirectional
variant of the dependently typed calculus of constructions. Languages with operator overloading may
provide interesting localization decisions. So that overloading can be resolved during typing and
marking, the modes of one or both operand must be changed to synthesis. If the operator is not
defined given the synthesized operand type(s), an error might be marked on the operator itself.

### 2023-10-25 related.tex
Fix more typos and address some of the feedback

Hazelnut is in turn rooted in gradual type theory \cite{Siek06a, siek2015refined}, and we make extensive use of (only) the static aspects of gradual typing, namely the universal consistency of the unknown type, to enable recovery from marked errors, which can leave missing type information.

@@

Type error localization is a well-studied problem in practice. This paper is the first to formally support the intuition that, in the words of \cite{BidirTyping}, ``bidirectional typing improves error locality''. Although there has been considerable folklore around error localization 
for systems with local type inference, the problem has received little formal attention. We hope that this paper, with its rigorous formulation
of type error localization and recovery for bidirectionally typed languages, will provide more rigorous grounding to language server development,
much as bidirectional typing has done for type checker development.

>>>>

Hazelnut is in-turn rooted in gradual type theory \cite{Siek06a, siek2015refined}, and we make extensive use of (only) the static aspects of gradual typing, namely the universal consistency of the unknown type, to enable recovery from marked errors, which can leave missing type information.

@@

Type error localization is a well-studied problem in practice. This paper is the first to formally support the intuition that, in the words of \citet{BidirTyping}, ``bidirectional typing improves error locality.'' Although there has been considerable folklore around error localization 
for systems with local type inference, the problem has received little formal attention. We hope that this paper, with its rigorous formulation
of type error localization and recovery for bidirectionally typed languages, will provide more rigorous grounding to language server development,
much as bidirectional typing has done for type checker development.
